Time and space specifications for stencils

### 2016-06-16 paper.tex
inference

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{tikz}

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,baselinestretch=0.97,numbers=left}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.97}

@@

% Writing macros
\newcommand{\ie}{\emph{i.e..}}
\newcommand{\eg}{\emph{e.g.}}

@@

\subsection{A semantic model for specifications}
\label{sec:semantics}

@@

@@

\section{Specification inference and checking}
\label{sec:analysis}

@@

In the following, we defined
\emph{base induction variables} to be the variables
 defined by a ``for'' loop (\eg{}, that get incremented each time the
loop is re-entered).

\begin{definition}[Neighbourhood indices]
An array subscript $e$ is a \emph{neighbourhood index} if,
for an induction variable $i$, then $e \equiv i + a$ or $e \equiv i - a$
where $a$ is a constant. The relation $\equiv$ identifies terms
up-to commutativity of $+$ and the inverse
relation of $+$ and $-$ (\eg{}, $(-b) + i \equiv i - b$).
In the following, we classify neighbourhood indices 
using the predicate $\textsf{neigh}$.
\end{definition}

\paragraph{Step 1: Data-access analysis}

\dnote{Might want to do 'stencil' (gather) and 'mould' (scatter)
  patterns separately}
\begin{align*}
\begin{array}{lll}
\interp{e_1 = e_2}    & = \interp{e_2} \\
\interp{e_1 e_2}      & = \interp{e_1} \cup \interp{e_2} \\
\interp{a(\tilde{e})} & = \{e | e \in \tilde{e}, \textsf{neigh}(e)\} & \textit{where each $e \in
                                          \tilde{e}$ is a constant or constant translation}
\end{array}
\end{align*}

Index analysis is a \emph{coeffect analysis}.

\paragraph{Step 2: Coalesce contiguous indices into regions}

% fairly mechanical

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{tikz}
\usepackage{multirow}

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.65cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.65cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98}

@@

% Writing macros
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}

@@

@@

\section{Semantics of specifications; a model}
\label{sec:semantics}

@@

\section{Inference, checking, and synthesis}
\label{sec:analysis}

@@

\noindent
We briefly sketch the algorithms for inferring and checking
specifications and for synthesising programs from specifications.
Checking and synthesis use the set-based semantic model of
Section~\ref{sec:semantics}.

% SYNTAX OPERATIONS AND PREDICATES
\newcommand{\neigh}{\textsf{neigh}}
\newcommand{\arrayTy}{\textsf{array}}
\newcommand{\rhsExp}{\textsf{rhsExp}}
\newcommand{\var}{\textsf{var}}

%% VECTOR NOTATIONS
\newcommand{\vect}[1]{\textbf{#1}}
\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}#1\\#2\end{array}\right]}
\newcommand{\stwo}[4]
%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
{\setlength{\arraycolsep}{0.1em}
\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}

%% OPERATIONS ON SPANS and VECTORS 
\newcommand{\containedin}{\sqsubseteq}

\paragraph{Basic syntax definitions} In the following, $v$
  ranges over variables, $s$ over statements, and $e$ over expressions
  (which may be impure).  By ``variable'', we mean in the imperative
  sense (\ie{}, named binders to memory cells) instead of mathematical
  variables. The function \rhsExp{} maps statements to a set of
  expressions which occur in right-hand positions (\ie{}, not the
  target of an assignment). The function \var{} maps expressions to a
  set of the variables used in right-hand positions.

\begin{definition}[Standard: base induction variable]
An integer variable \textit{i} is a \emph{base induction variable} 
within the scope of a ``for'' loop (or equivalent for the target
language, \eg{}, \texttt{do} in Fortan for our implementation) if it
is the control variable of the loop, incremented by $1$ per
iteration. 
\end{definition}

\begin{definition}[Standard: array subscript]
An \emph{array subscript} is an expression index an array, which in
our abstract we write as $v(\bar{e})$ where $\bar{e}$ is shorthand for
a syntactic list of indexing expressions. 
\end{definition}

\begin{definition}[Neighbourhood index]
An array subscript $e$ is a \emph{neighbourhood index} if,
for an induction variable $i$, then $e \equiv i + a$ or $e \equiv i - a$
where $a$ is a constant of integer type. The relation $\equiv$ identifies terms
up-to commutativity of $+$ and the inverse
relation of $+$ and $-$ (\eg{}, $(-b) + i \equiv i - b$).
In the following, we classify neighbourhood indices 
using the predicate \neigh{}.
\end{definition}

\subsection{Inference}

We demonstate the main steps of the inference procedure over the
following example program which computes the mean value 
of a five-point stencil at each index of the input array:
%%
\begin{ExmVerbatim}
do i = 1, (n-1)
   do j = 1, (m-1)
      x       = a(i-1, j) + a(i+1, j) 
      y       = a(i, j-1) + a(i, j+1)
      b(i, j) = (a(i, j) + x + y) / 5.0
   end do
end do
\end{ExmVerbatim}
%%
\paragraph{Step 1: Standard control and dataflow analyses}
The inference relies on some standard program analyses, computed
before the main inference procedure:
%
\begin{enumerate}
\item basic blocks (CFG);
\item induction variables per basic block;
\item (interprocedural) dataflow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,  
  mapping an expression to the set of all expressions
  with forwards dataflow to this expression, based on the transitive
  closure of the flows graph);
\item type information per variable, where we use the predicate
\arrayTy{} to classify variables of array type. 
\end{enumerate}
%

\paragraph{Step 2: Data-access analysis}

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statment. More formally, a function
$\textsf{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\begin{array}{lr}
\textsf{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{convert}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where each
index indexing expression in $\bar{e}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed from each array variable
in this flow set to a representation of the subscripts, computed
with \textsf{convert}.

The \textsf{convert} function maps subscripts to a column vector
representation with values drawn from $\mathbb{Z}_\top = \mathbb{Z} \cup \{\top\}$
where $\mathbb{Z}$ represents the offset of a neighbourhood index
and $\top$ represents any non-neighbourhood subscripts, defined:
%
\begin{align*}
& \textsf{convert}((e_1, \ldots, e_n)) = [\textsf{conv}(e_1), \ldots,
  \textsf{conv}(e_n)] \\
& \textsf{conv}(e) \begin{cases}
\pm a & \neigh(e) \wedge \, e \equiv i \pm a \\
\top & \textit{otherwise} 
\end{cases} 
\end{align*}
%
For our example, the \textsf{analyse} function matches on
line 5, with the following set for $\textsf{flowTo}(a(i, j) + x +
  y)$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textsf{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwo{-1}{0}, \vtwo{1}{0},
          \vtwo{0}{-1}, \vtwo{0}{1}, \vtwo{0}{0}\}
\end{align*}
%

\paragraph{Step 3: Coalesce contiguous indices into regions}

Let $M$ range over the finite maps generated by \textsf{analyse}.  For
each $v \in \mathsf{dom}(M)$, the next step of the algorithm coalesces
the set of relative index vectors $M(v)$ into contiguous regions within an
$n$-dimensional space.

Let $\vect{x}, \vect{y}, \vect{z}$ range over the column vectors of size
$n$ whose values are drawn from $\mathbb{Z}_\top$.
We write $\vect{x}_i$ for the $i$-th value of vector $\vect{x}$ (1
indexed) and $\vect{x}_{i:j}$ for the subvector from entry $i$ to
entry $j$ (inclusive).

\begin{definition}[Span]
  A \emph{span} represents an $n$-dimensional box (\emph{hyperrectangle}) as
  a pair of $n$-dimensional vectors (represented as a $2 \times n$
  matrix) giving the co-ordinates of the lower-bound vertex (first
  column) and the upper-bound vertex (second column). We use the
  notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
  the lower and upper bound vecvtors. 
\end{definition}

Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \; 
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then compute for the following
procedure \textsf{regions} which, for per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional spaace. That is, we compute
 $(\mu \; \textsf{regions}) N$ where \textsf{regions}
is defined as follows for each variable in the map:
%
\begin{enumerate}

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$, we pair the
permutation function with the set of permuted spans so that 
the spans can be unpermuted later. 
%
\[
P(v) = \bigcup_{i \in !n} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) = 
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0} 
\stwo{-1}{0}{-1}{0} 
\stwo{0}{1}{0}{1} 
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{ 
 \stwo{0}{0}{0}{0}
 \stwo{0}{1}{0}{1}
 \stwo{0}{-1}{0}{-1}
 \stwo{1}{0}{1}{0}
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.

\item Sort each permutation set into an ordered list, based on the ordering:
\[
\vect{x} \leq \vect{y} = \exists i \, . \; \vect{x}_i \leq \vect{y}_i 
\; \wedge \; \vect{x}_{i:n} = \vect{y}_{i:n}
\]
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0} 
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0} 
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial $\bullet$ which 
coalescees and contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
(\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) = (\vect{y}^L_{1:n}, \vect{y}^U_{1:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0}, 
\stwo{0}{1}{0}{1}, 
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Unpermute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{y}^L, \pi \vect{y}^U]
 \mid [\vect{y}^L, \vect{y}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) = 
\{\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0}
\stwo{0}{1}{0}{1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{-1}{0}{1}
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Filter by $\containedin$, defined: 
%
\begin{align*}
(\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = 
\vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
  (\vect{y}^L_{1:n}, \vect{y}^U_{1:n})
\end{align*}

Yields:
\begin{align*}
\stwo{-1}{0}{1}{0}, 
\stwo{0}{-1}{0}{1}
\end{align*}
\end{enumerate}

\begin{tikzpicture}
\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\draw[->] (0.5,2) -- (2,2);
\node at (1.25,1.22) {i, j};
\draw[thick] (1,1) rectangle (1.5,1.5);
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
%
\begin{tikzpicture}
\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\draw[->] (0.5,2) -- (2,2);
\node at (1.25,1.22) {i, j};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
%
\begin{tikzpicture}
\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\draw[->] (0.5,2) -- (0.5,0.5);
\node at (1.25,1.22) {i, j};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
